---
title: "Analysis"
---

## Key takeaways

- **Topic:** July **2024** **street-animal feeding** operations across **Ankara ilçeler**, from health-directorate-style operational records.
- **Data:** **367** neighborhood-day records with feeding-point counts and **kg/day** dry food averages, plus vehicle and personnel fields.
- **Important aspects:** **District names** needed trimming; **food** was text-encoded; **vehicle** column mixes numbers, `"-"`, and a **`Y3`** typo.
- **Interesting results:** **Çankaya**, **Keçiören**, and **Mamak** dominate row counts; **kg/day per record** is right-skewed with a long tail of large feeding averages.
- **Main outcome:** Clear **district comparisons** and **temporal coverage** (26 distinct calendar days in July) support a discussion of **resource concentration** and **data quality** for policy narratives.

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(scales)
library(lubridate)
library(here)

load(here::here("data", "feeding_clean.RData"))
```

## Coverage in time and space

```{r}
#| label: tbl-summary

feeding |>
  summarise(
    rows = n(),
    first_day = min(tarih, na.rm = TRUE),
    last_day = max(tarih, na.rm = TRUE),
    distinct_days = n_distinct(as.Date(tarih)),
    distinct_ilce = n_distinct(ilce),
    missing_mama = sum(is.na(mama_kg)),
    missing_arac = sum(is.na(arac_sayisi)),
    missing_personel = sum(is.na(personel_sayisi))
  ) |>
  knitr::kable(caption = "Dataset coverage and missingness")
```

The source table does not always include **every calendar day** in July (`distinct_days` \< 31). That is a **reporting / data collection** feature to mention in your presentation, not something we impute.

## District activity (row counts)

```{r}
#| label: fig-rows-ilce
#| fig-cap: "Number of neighborhood-day records per district (ilçe)."

ilce_counts <- feeding |>
  count(ilce, name = "rows") |>
  mutate(ilce = stats::reorder(factor(ilce), rows))

ggplot(ilce_counts, aes(x = ilce, y = rows)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(x = NULL, y = "Rows (mahalle × day)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## Distribution of dry food (kg/day)

```{r}
#| label: fig-mama-hist
#| fig-cap: "Histogram of parsed average dry food (kg/day) per record."

ggplot(feeding, aes(x = mama_kg)) +
  geom_histogram(fill = "#A23B72", color = "white", bins = 30, boundary = 0) +
  labs(x = "kg/day (parsed)", y = "Count") +
  theme_minimal(base_size = 12)
```

## District comparison of kg/day

```{r}
#| label: fig-mama-ilce
#| fig-cap: "Distribution of kg/day by district for districts with at least 15 records."

big_ilce <- feeding |>
  count(ilce) |>
  filter(n >= 15) |>
  pull(ilce)

feeding_sub <- feeding |>
  filter(ilce %in% big_ilce) |>
  mutate(ilce = stats::reorder(factor(ilce), mama_kg, FUN = median, na.rm = TRUE))

ggplot(feeding_sub, aes(x = ilce, y = mama_kg)) +
  geom_boxplot(outlier.alpha = 0.35, fill = "#F18F01") +
  coord_flip() +
  scale_y_continuous(labels = label_number(suffix = " kg")) +
  labs(x = NULL, y = "Average dry food (kg/day)") +
  theme_minimal(base_size = 12)
```

## Vehicles and personnel

These columns were **mostly constant** in the July extract (many rows share the same small integers), with **`"-"` coded gaps**. Your write-up should interpret them cautiously: they may reflect **fleet / staffing assumptions** in reporting rather than precise daily counts.

```{r}
#| label: tbl-staff

feeding |>
  summarise(
    arac_distinct = n_distinct(arac_sayisi, na.rm = TRUE),
    personel_distinct = n_distinct(personel_sayisi, na.rm = TRUE),
    share_arac_missing = mean(is.na(arac_sayisi)),
    share_personel_missing = mean(is.na(personel_sayisi))
  ) |>
  knitr::kable(digits = 3)
```

## Observations per day

```{r}
#| label: fig-by-day
#| fig-cap: "Number of records per calendar day."

daily <- feeding |>
  mutate(day = as.Date(tarih)) |>
  count(day, name = "rows")

ggplot(daily, aes(x = day, y = rows)) +
  geom_line(color = "#C73E1D", linewidth = 0.8) +
  geom_point(color = "#C73E1D", size = 2) +
  scale_x_date(date_labels = "%d %b", date_breaks = "3 days") +
  labs(x = NULL, y = "Rows per day") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Conclusions (for your narrative)

- **Spatial concentration:** A few districts account for most neighborhood-day rows — tie this to population, urban form, or program emphasis **only as hypotheses** unless you merge external data.
- **Skewed feeding amounts:** A long right tail suggests a mix of **small local drops** and **larger consolidated deliveries**; outliers deserve qualitative mention.
- **Operational fields:** Vehicle/personnel columns are **sparse / low-variance**; do not over-claim precision.

## AI use disclosure (if applicable)

If you extend this analysis with AI-generated text, **label it** and add the **prompt in a footnote**; validate all numeric statements against the `feeding` object.
