---
title: "Analysis"
---

## Key takeaways

- **Topic:** **Street-animal feeding** operations across **Ankara districts (`ilçeler`)** from **July 2024 to June 2025**, based on monthly health-directorate operational records.
- **Data:** ~3,700+ neighborhood-day records aggregated from **11 monthly workbooks**, with feeding-point counts, **kg/day** of dry food, and vehicle / personnel staffing.
- **Important aspects:** District names needed trimming; food / dates / staff fields were **text-encoded** with inconsistent formats (March 2025 dates as `"1.03.2025"`, vehicles as `"-"` or the typo `"Y3"`).
- **Interesting results:** A handful of districts (notably **Çankaya**, **Keçiören**, **Mamak**) dominate row counts; **monthly totals fluctuate**, and the kg-per-record distribution is **right-skewed** with consolidated large drops in the tail.
- **Main outcome:** A reproducible **Quarto + R** pipeline plus a publicly downloadable cleaned `.RData` enabling year-scale district comparisons and discussion of **resource concentration and data quality** in municipal animal-welfare operations.

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(scales)
library(lubridate)
library(here)

load(here::here("data", "feeding_clean.RData"))
```

## Coverage in time and space

```{r}
#| label: tbl-summary

feeding |>
  summarise(
    rows               = n(),
    months             = n_distinct(yil_ay),
    first_day          = min(tarih, na.rm = TRUE),
    last_day           = max(tarih, na.rm = TRUE),
    distinct_days      = n_distinct(as.Date(tarih)),
    distinct_ilce      = n_distinct(ilce),
    missing_mama       = sum(is.na(mama_kg)),
    missing_arac       = sum(is.na(arac_sayisi)),
    missing_personel   = sum(is.na(personel_sayisi))
  ) |>
  knitr::kable(caption = "Dataset coverage and missingness across all months")
```

The monthly workbooks **do not always report every calendar day**, so `distinct_days` is well below the calendar maximum. We treat that as a reporting feature rather than imputing days.

## Monthly volume

How many records per month, and how much dry food (kg) was reported per month?

```{r}
#| label: fig-monthly
#| fig-cap: "Records per month and total reported kg of dry food per month."
#| fig-height: 6

monthly <- feeding |>
  filter(!is.na(tarih)) |>
  mutate(month_start = lubridate::floor_date(tarih, "month")) |>
  group_by(month_start) |>
  summarise(
    rows = n(),
    total_kg = sum(mama_kg, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(month_start)

p1 <- ggplot(monthly, aes(month_start, rows)) +
  geom_col(fill = "#2E86AB") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(x = NULL, y = "Records per month") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p2 <- ggplot(monthly, aes(month_start, total_kg)) +
  geom_col(fill = "#A23B72") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = label_number(suffix = " kg", big.mark = ",")) +
  labs(x = NULL, y = "Reported dry food (kg / month)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p1
p2
```

Two complementary views:

- The **record count** reflects how many neighborhood-days were filed that month — partly a function of reporting effort.
- The **kg total** reflects the actual feeding workload (after parsing strings like `"84 KG"`).

## District activity (row counts)

```{r}
#| label: fig-rows-ilce
#| fig-cap: "Number of neighborhood-day records per district across the full year."

ilce_counts <- feeding |>
  count(ilce, name = "rows") |>
  mutate(ilce = stats::reorder(factor(ilce), rows))

ggplot(ilce_counts, aes(x = ilce, y = rows)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(x = NULL, y = "Rows (mahalle × day, all months)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## Distribution of dry food (kg/day)

```{r}
#| label: fig-mama-hist
#| fig-cap: "Histogram of parsed daily dry food per record (kg)."

ggplot(feeding |> filter(!is.na(mama_kg)), aes(x = mama_kg)) +
  geom_histogram(fill = "#A23B72", color = "white", bins = 40, boundary = 0) +
  scale_x_continuous(labels = label_number(suffix = " kg")) +
  labs(x = "kg/day (parsed)", y = "Count") +
  theme_minimal(base_size = 12)
```

The distribution is **right-skewed**: most records are modest (low single-digit to tens of kg), but a long tail of larger consolidated deliveries pulls the mean up.

## District comparison of kg/day

```{r}
#| label: fig-mama-ilce
#| fig-cap: "Distribution of kg/day by district (districts with ≥ 50 records across the year)."

big_ilce <- feeding |>
  count(ilce) |>
  filter(n >= 50) |>
  pull(ilce)

feeding_sub <- feeding |>
  filter(ilce %in% big_ilce, !is.na(mama_kg)) |>
  mutate(ilce = stats::reorder(factor(ilce), mama_kg, FUN = median, na.rm = TRUE))

ggplot(feeding_sub, aes(x = ilce, y = mama_kg)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#F18F01") +
  coord_flip() +
  scale_y_continuous(labels = label_number(suffix = " kg")) +
  labs(x = NULL, y = "Average dry food (kg/day)") +
  theme_minimal(base_size = 12)
```

## Top districts over time

How does the **monthly kg total** vary across the top five districts (by record count)?

```{r}
#| label: fig-ilce-time
#| fig-cap: "Monthly total kg of dry food for the five most-reported districts."
#| fig-height: 5

top5 <- feeding |>
  count(ilce, sort = TRUE) |>
  slice_head(n = 5) |>
  pull(ilce)

feeding_top <- feeding |>
  filter(ilce %in% top5, !is.na(tarih)) |>
  mutate(month_start = lubridate::floor_date(tarih, "month")) |>
  group_by(ilce, month_start) |>
  summarise(total_kg = sum(mama_kg, na.rm = TRUE), .groups = "drop")

ggplot(feeding_top, aes(month_start, total_kg, color = ilce)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = label_number(suffix = " kg", big.mark = ",")) +
  labs(x = NULL, y = "Total reported kg / month", color = "İlçe") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "bottom"
  )
```

## Vehicles and personnel

These columns are **low-variance** (often constant within a month) and frequently use `"-"` for missing values, so they describe *fleet / staffing assumptions* more than precise daily counts.

```{r}
#| label: tbl-staff

feeding |>
  summarise(
    arac_distinct          = n_distinct(arac_sayisi, na.rm = TRUE),
    personel_distinct      = n_distinct(personel_sayisi, na.rm = TRUE),
    share_arac_missing     = mean(is.na(arac_sayisi)),
    share_personel_missing = mean(is.na(personel_sayisi))
  ) |>
  knitr::kable(digits = 3, caption = "Vehicle and personnel coverage")
```

## Observations per day

```{r}
#| label: fig-by-day
#| fig-cap: "Number of records per calendar day across the full reporting window."

daily <- feeding |>
  filter(!is.na(tarih)) |>
  mutate(day = as.Date(tarih)) |>
  count(day, name = "rows")

ggplot(daily, aes(x = day, y = rows)) +
  geom_line(color = "#C73E1D", linewidth = 0.5, alpha = 0.6) +
  geom_point(color = "#C73E1D", size = 1.2, alpha = 0.7) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(x = NULL, y = "Rows per day") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Conclusions

- **Spatial concentration:** A handful of districts (Çankaya, Keçiören, Mamak and a few others) account for most of the reported activity. Tying this to population, urban form, or program emphasis remains a **hypothesis** without external data.
- **Temporal patterns:** Monthly volume is uneven, partly driven by reporting effort rather than real demand. The kg totals are less sensitive to that and offer a more reliable workload proxy.
- **Right-skewed food amounts:** A long upper tail represents **consolidated bulk feedings** at specific points; the median is a better summary than the mean.
- **Operational fields:** Vehicle/personnel are too coarse to claim per-day precision — interpret as background fleet/staffing context.
- **Data quality matters:** Different months use different date formats, and one row carries an obvious 2022 typo — a reminder that reproducible cleaning is part of the analysis, not a side concern.

## AI use disclosure

If you extend this analysis with AI-generated text, **label that section** and include the **prompt as a footnote**; validate every numeric statement against the `feeding` object.
